A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-05 tex/codatadp.tex
Minor cleanup, starting solution soundness.

\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\ec}[1]{ [ #1 ] }

% change?
\newcommand{\thD}{T_{D}}

@@

\rem{Discuss signature $\Sig$, theory $\thD$}
This section presents a calculus for determining the satisfiability of conjunctions of $\thD$-constraints.

@@

To explain the rules in Figure~\ref{fig:ab-rules}, we introduce the following notions.
First, an edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is a transition between two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
(which we will call its \emph{source} and \emph{destination} respectively),
labeled by a constructor $C$ and integer $n$.
We say that edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is \emph{induced by $\Ec$} if and only if $C( \vec u ) \in \ec{t_1}$ for some $\vec u$ and $\ec{t_2} = \ec{u_n}$.
In other words, $t_1$ is equivalent to a $C$-application whose $n^{th}$ argument is equivalent to $t_2$.
A \emph{path} $p$ is a non-empty ordered list of edges $e_1, \ldots e_n$ such that the destination of $e_i$ is the source of $e_{i+1}$ for $i = 1, \ldots (n-1)$,
and the sources of $e_1, \ldots, e_n$ are distinct.
We say that path $p$ is \emph{induced by $\Ec$} if and only if each of its edges are induced by $\Ec$.
Notice that since $\Ec$ contains a finite set of equivalence classes, all paths induced by $\Ec$ have finite length.
We call the \emph{source} of a path is the source of its first edge, and the \emph{destination} of a path is the destination of its last edge.
We say a path $p$ is \emph{maximal} if its destination $\ec{t}$ is either
the source of an edge in $p$, 
or does not contain an application of a constructor (including the case when $t$ is not of datatype/codatatype type).
The set $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ is the smallest set containing all maximal paths with source $\ec{t_1}$ and destination $\ec{t_2}$ that are induced by $\Ec$.
The set $\tpath{\Ec}{\ec{t_1}}$ is the union of $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ for all $t_2 \in \tEc$.
We will write, for instance, $\ec{t_1} \rightarrow^{C_1}_{n_1} \ec{t_2} \rightarrow^{C_2}_{n_2} \ec{t_3}$ 
as shorthand for the path containing an edge from $\ec{t_1}$ to $\ec{t_2}$, followed by one from $\ec{t_2}$ to $\ec{t_3}$.

@@

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S$ be the set of terms occurring as the argument of a selector in $\tEc$.
For each term $t \in \tEc$ of finite datatype type $\tau$,
let $S^0_t$ be the set $\{ t \}$,
and for $i \geq 0$ let $S^{i+1}_t =$ $S^i_t \cup \{ s( u ) \mid u : \upsilon \in S^i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$.
Let $S^\ast_t$ be the fixed point of this sequence for all $t$, which we know is finite since the values of $\tau$ are of finite size.
Let $S^\ast$ be the union of $S^\ast_t$ for all such $t$,
and let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t : \tau \in S \cup S^\ast, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node $\Ec'$ is such that 
$\mathcal{T}(\Ec') \subseteq \tcEc $, and moreover thus is the result of adding an equality between two terms from $\tcEc$ to its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

@@

\begin{lemma}[Solution Soundness]
If there exists a derivation tree containing a saturated node with root node $\Ec$, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
TODO
\end{proof}

>>>>

\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}

% change?
\newcommand{\thO}{T_{o}}
\newcommand{\thD}{T_{d}}

@@

\rem{Discuss signature $\Sig$, theory $\thD$}
\rem{DPLL(T) : we can restrict ourselves to conjunctions.}
This section presents a calculus for determining the satisfiability of conjunctions of $\thD$-constraints.

@@

To explain the rules in Figure~\ref{fig:ab-rules}, we introduce the following notions.
First, an edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is a transition between two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
(which we will call its \emph{source} and \emph{destination} respectively),
labeled by a constructor $C$ and integer $n$.
We say that edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is \emph{induced by $\Ec$} if and only if $C( \vec u ) \in \ec{t_1}$ for some $\vec u$ and $\ec{t_2} = \ec{u_n}$.
In other words, $t_1$ is equivalent to a $C$-application whose $n^{th}$ argument is equivalent to $t_2$.
A \emph{path} $p$ is a non-empty ordered list of edges $e_1, \ldots e_n$ such that the destination of $e_i$ is the source of $e_{i+1}$ for $i = 1, \ldots (n-1)$,
and the sources of $e_1, \ldots, e_n$ are distinct.
We say that path $p$ is \emph{induced by $\Ec$} if and only if each of its edges are induced by $\Ec$.
Notice that since $\Ec$ contains a finite set of equivalence classes, all paths induced by $\Ec$ have finite length.
We call the \emph{source} of a path is the source of its first edge, and the \emph{destination} of a path is the destination of its last edge.
We say a path $p$ is \emph{maximal} if its destination $\ec{t}$ is either
the source of an edge in $p$, 
or does not contain an application of a non-nullary constructor (including the case when $t$ is not of datatype/codatatype type).
The set $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ is the smallest set containing all maximal paths with source $\ec{t_1}$ and destination $\ec{t_2}$ that are induced by $\Ec$.
The set $\tpath{\Ec}{\ec{t_1}}$ is the union of $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ for all $t_2 \in \tEc$.
We will write, for instance, $\ec{t_1} \rightarrow^{C_1}_{n_1} \ec{t_2} \rightarrow^{C_2}_{n_2} \ec{t_3}$ 
as shorthand for the path containing an edge from $\ec{t_1}$ to $\ec{t_2}$, followed by one from $\ec{t_2}$ to $\ec{t_3}$.

@@

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S$ be the set of terms occurring as the argument of a selector in $\tEc$.
Let $T$ be the set of terms in $\tEc$ of finite datatype type.
For each term $t \in T$,
let $S^0_t$ be the set $\{ t \}$,
for $i \geq 0$ let $S^{i+1}_t =$ $S^i_t \cup \{ s( u ) \mid u : \upsilon \in S^i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$,
and let $S^\ast_t$ be the fixed point of this sequence.
We know this is a finite set for each $t$ since the values of $t$'s type are of finite size.
Let $S^\ast$ be the union of $S^\ast_t$ for all $t \in T$,
and let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t : \tau \in S \cup S^\ast, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node $\Fc$ is such that 
$\mathcal{T}(\Fc) \subseteq \tcEc $, and moreover thus contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

@@

\begin{lemma}[Solution Soundness]
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
%First, we construct a valuation $\Val$ as follows.
Let $C$ be the set of variables occurring in the same equivalence class as a constructor term,
and let $X$ be the set of variables occurring in $\mathcal{T}(\Fc)$ but not in $C$.

\qed
\end{proof}
